Parallel Simulation of Adaptive Random Boolean Networks

### 2016-07-14 Paper/paper.tex
Fix surnames, fix attractor lengths of modular networks
Add description of Knuth algorithm

% Authors are joined by \and. Their affiliations are given by \inst, which indexes into the list
% defined using \institute
%
\author{
	Kirill Kuvshinov\inst{1}
	\and
	Klavdiya Bochenina\inst{1}
	\and
	Piotr Gorsky\inst{2}
	\and
	Janusz Holyst\inst{1}\inst{2}
}

@@

	As RBNs are applied to model biological structures, it is interesting to study the evolution of RBNs. Core of the evolution process is a constant searching for \footnote{Tu sie nie zgadzam, ze powinno być of. Istnieja 2 zwroty: 'search for' oraz 'in search of'. Ten drugi znaczy 'w poszukiwaniu'.} the most optimal configuration \cite{aldana2003boolean,sole1996extinction,zimmermann2004coevolution}. 
	An RBN modification that incorporates system evolution is {\it adaptive} RBN (ARBN) \cite{mlb,soc,haruna14relationship,Gorski2016}. 
	In RBNs nodes can be divided into two groups of active and frozen nodes. Nodes of the former group change their state during network attractor. On the other hand, the frozen nodes keep their state. 
	The basis of ARBN is a natural idea that overactive nodes should be quietened down and underactive (frozen) ones should be stimulated. This concept is included in an activity-dependent rewiring rule (ADRR) \cite{bornholdt2000topological}, which is as follows: 
	if the node’s mean state $<\sigma_i>$ during the network attractor is $0$ or $1$, then this node
	is considered to be frozen and one new incoming edge is
	added to this node. Otherwise, this node is considered to be active and one of
	its incoming edges is randomly chosen and deleted.

@@

	\paragraph{Knuth algorithm.} (The description of Knuth algorithm)

@@

	Intuitively, Liu-Bassler algorithm of attractor's search would have shown better performance than the Knuth's one, as the latter requires at least $3 \mu$ state updates to find an attractor of length $\mu$. But in order to use Liu-Bassler algorithm one needs to choose the checkpoints correctly, which requires an a-priori knowledge of attractor and transient period distributions. This distribution depends on a network size, connectivity and structure (e. g. modular networks tend to have longer (shorter?) attractors than non-modular ones [ref]). We obtained an empirical distribution of attractor and transient lengths for different kinds of networks in our preliminary experiments with Knuth's algorithm. The checkpoints $T=[100, 200, 1000, 2000, ... \mu_{max}, 2 \mu_{max}]$ seem to be the best trade-off between the number of state updates and the percentage of attractors found. However, with these checkpoints the difference in the number of state updates between the two algorithms is negligible. We used Knuth's algorithm in our future experiments, as it is more general and requires no a-priori knowledge of the network's behaviour.

>>>>

% Authors are joined by \and. Their affiliations are given by \inst, which indexes into the list
% defined using \institute
%
\author{
	Kirill Kuvshinov\inst{1}
	\and
	Klavdiya Bochenina\inst{1}
	\and
	Piotr G{\'{o}}rski\inst{2}
	\and
	Janusz Ho{\l}yst\inst{1}\inst{2}
}

@@

	As RBNs are applied to model biological structures, it is interesting to study the evolution of RBNs. Core of the evolution process is a constant searching for \footnote{Tu sie nie zgadzam, ze powinno być of. Istnieja 2 zwroty: 'search for' oraz 'in search of'. Ten drugi znaczy 'w poszukiwaniu'.} the most optimal configuration \cite{aldana2003boolean,sole1996extinction,zimmermann2004coevolution}. 
	An RBN modification that incorporates system evolution is {\it adaptive} RBN (ARBN) \cite{mlb,soc,haruna14relationship,Gorski2016}. 
	In RBNs nodes can be divided into two groups of active and frozen nodes. Nodes of the former group change their state during network attractor. On the other hand, the frozen nodes keep their state. 
	The basis of ARBN is a natural idea that overactive nodes should be quietened down and underactive (frozen) ones should be stimulated. This concept is included in an activity-dependent rewiring rule (ADRR) \cite{bornholdt2000topological}, which is as follows: 
	if the node’s mean state $\langle\sigma_i\rangle$ during the network attractor is $0$ or $1$, then this node
	is considered to be frozen and one new incoming edge is
	added to this node. Otherwise, this node is considered to be active and one of
	its incoming edges is randomly chosen and deleted.

@@

	\paragraph{Knuth algorithm.} The algorithm is based on the following observation: $\mathbf{S}(\tau) = \mathbf{S}(2\tau)$ if and only if $\tau$ is a multiple of an attractor length $\mu$, i. e. $\tau = n\mu$. In order to find $\tau$, two copies of a Boolean network are launched, one copy performs one state update per iteration, and the other one performes two. Then these states are compared. Nodes' mean state on an attractor, which is required to apply ADRR, can be found without determining the value of $\mu$. First, one should find sum of nodes' states on $t \in (\tau, 2\tau]$ by calulating state sum vector $\Sigma = \mathbf{S}(2\tau) + \mathbf{S}(2\tau + 1) - \mathbf{S}(\tau)$ on each iteration. After $\tau$ is found, mean state of the $i$-th node on the attractor is $\langle\sigma_i\rangle = \Sigma_i / \tau$. Having found $\tau$ it is possible to obtain the length of the attractor by finding $\mu$ such that $\mathbf{S}(\tau) = \mathbf{S}(\tau + \mu)$.

@@

	Intuitively, Liu-Bassler algorithm of attractor's search would have shown better performance than the Knuth's one, as the latter requires at least $3 \mu$ state updates to find an attractor of length $\mu$. But in order to use Liu-Bassler algorithm one needs to choose the checkpoints correctly, which requires an a-priori knowledge of attractor and transient period distributions. This distribution depends on a network size, connectivity and structure (e. g. modular networks tend to have shorter attractors than non-modular ones \cite{Gorski2016}). We obtained an empirical distribution of attractor and transient lengths for different kinds of networks in our preliminary experiments with Knuth's algorithm. The checkpoints $T=[100, 200, 1000, 2000, ... \mu_{max}, 2 \mu_{max}]$ seem to be the best trade-off between the number of state updates and the percentage of attractors found. However, with these checkpoints the difference in the number of state updates between the two algorithms is negligible. We used Knuth's algorithm in our future experiments, as it is more general and requires no a-priori knowledge of the network's behaviour.
